---
title: "Matrix inversion accuracy"
subtitle: ""
summary: "Method for diagnosing the accuracy of digital matrix inversion."
authors: []
tags: []
categories: []
date: 2023-09-13
featured: false
draft: false

# Featured image
# To use, add an image named `featured.jpg/png` to your page's folder.
# Focal points: Smart, Center, TopLeft, Top, TopRight, Left, Right, BottomLeft, Bottom, BottomRight.
image:
  caption: ""
  focal_point: ""
  preview_only: false
---

# Theory

Inversion of large matrices is difficult.
Newman (1974) proposes a method for testing inversion accuracy.

Let $A$ be the (square, non-singular) matrix to be inverted.
$X$ is the approximation of $A^{-1}$ obtained by calculation.
$N(Y)={sqrt{sum_{i,j}{y_{i,j}^2}}$ is the Euclidean norm of a matrix $Y=(y_{i,j})$.

$R = I - AX$ is the difference (ideally zero) between the identity matrix and the product of $A$ and its calculated inverse.

The norm of $R$ is not a good indicator of inversion accuracy (Newman, 1974, section 3).
On the other hand, the norm of $A^{-1} - X$ is, but $A^{-1}$ is unknown.
Newman shows that
$$\frac{N(XR)}{1 + N(XR)} \leq N(A^{-1} - X) \leq \frac{N(XR)}{1 - N(XR)},$$
which allows to bound the value of $N(A^{-1} - X)$ between the calculated values of $X$ and $R$.

The aim is for this value to be small, ideally zero if $A^{-1} = X$, relative to the norm of $X$.


# Example

The `size` parameter sets the size (number of rows and columns) of the matrices used, here a 1000 x 1000 matrix.

```{r}
size <- 1000
```

A matrix is created:
```{r}
A <- matrix(rnorm(size^2), nrow = size)
```

It is inverted numerically ($X$ is calculated), then $R$ and $XR$ are calculated:
```{r}
X <- solve(A)
AX <- A %*% X
R <- diag(size) - AX
XR <- X %*% R
```

The Euclidean norm function is defined:
```{r}
norm.matrix <- function(X) {
  sqrt(sum(X^2))
}
```

Matrix norms are calculated:
```{r}
# Matrix to invert
norm.matrix(A)
# Numerical inversion
norm.matrix(X)
# Inversion deviation
norm_R <- norm.matrix(R)
norm_XR <- norm.matrix(XR)
```

Finally, the bounds of the norm of $A^{-1} - X$ is provided:
```{r}
norm_XR / (1 + norm_R)
norm_XR / (1 - norm_R)
```

The norm of $A^{-1} - X$ is very close to 0: matrix inversion in R works well.


# Reference

Newman, M. (1974). How to Determine the Accuracy of the Output of a Matrix Inversion Program. *JOURNAL OF RESEARCH of the National Bureau of Standards - B. Mathematical Sciences*, [788(2), 65–68](https://nvlpubs.nist.gov/nistpubs/jres/78B/jresv78Bn2p65_A1b.pdf).
